Red Red Wine
========================================================

```{r init, include=FALSE}
knitr::opts_chunk$set(fig.width=9, fig.height=6, echo=FALSE, 
                      warning=FALSE, message=FALSE)

# Load libraries
library(corrplot)
library(ggplot2)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityReds.csv')
```

# Data Source
This is a data analysis of red wine samples, captured by  Paulo Cortez, 
Antonio Cerdeira, Fernando Almeida, Telmo Matos and Jose Reis.  Red Wines 
were analyzed on their physical qualities (chlorides, pH, alcohol by volume, 
etc...), and subjectively measured by wine experts.

# Univariate Plots

```{r echo=FALSE, Univariate_Plots}
# print dataset stats and summary information
dim(wine)

str(wine)
```

This dataset features 12 qualities of 1599 different red wines.  A thirteenth 
variable, 'X', is also listed, but it appears to be an index number for each 
wine.  All values within this dataset are numerical.

```{r}
# setting 'X' as an index and dropping the variable
rownames(wine) <- wine$X
wine <- subset(wine, select = -X)
```

```{r}
# plot histograms/frequency charts of each feature 
ggplot(data=wine, aes(x=as.factor(quality))) +
  geom_bar()

summary(wine$quality)
```

The quality scale was from 0 to 10, 0 being the worst and 10 being the best.  
The majority of the wines were given a score of 5 or 6.  Only a few wines were
given a score of 3, with none scoring lower.  A few wines were also given a 
score of 8, with none scoring higher.  No wines were given an exceptional or 
terrible rating, leading to this chart reflecting a normal distribution of 
quality.

Lets take a look at the physical properties of these wines.

```{r}
ggplot(data=wine, aes(x=fixed.acidity)) +
  geom_histogram(binwidth = .1) +
  scale_x_continuous(breaks = seq(4, 17, 1))

summary(wine$fixed.acidity)
```
For fixed acidity, the IQR is 2.1, however the range is 11.3.  In this 
positively skewed histogram, there are a lot of outliers over 14.

```{r}
ggplot(data=wine, aes(x=volatile.acidity)) +
  geom_histogram(binwidth = .01)+
  scale_x_continuous(breaks = seq(0, 2, .1))
summary(wine$volatile.acidity)
```

The histogram for volatile acidity and summary statistics also shows a 
positive skew, however there appear to be less outliers compared to fixed 
acidity.

```{r}
ggplot(data=wine, aes(x=citric.acid)) +
  geom_freqpoly(binwidth = .01)+
  scale_x_continuous(breaks = seq(0, 2, .1))
summary(wine$citric.acid)
```

All of the wines have some fixed acidity and volatile acidity.  However over 
100 of the samples contain no citric acid.  The citric acid  frequency polygon 
indicates a multi-modal distribution, with peaks around 0, 0.25, and 0.5 g/dm^3. 

```{r}
# create total.acid variable which sums of the values of all the acids
wine$total.acid <- wine$fixed.acidity + wine$volatile.acidity + wine$citric.acid

ggplot(data=wine, aes(x=total.acid)) +
  geom_histogram(binwidth = .1) +
  scale_x_continuous(breaks = seq(5, 18, 1))

summary(wine$total.acid)
```

The values of the three types of acid were summed up under a new variable: 
total acid.

```{r}
ggplot(data=wine, aes(x=free.sulfur.dioxide)) +
  geom_histogram(binwidth = 1)+
  scale_x_continuous(breaks = seq(0, 75, 5))
summary(wine$free.sulfur.dioxide)
```

This histogram for free sulfur dioxide is positively skewed with a long tail. 
As the 3rd quartile is 21 and the max is 72, there is a large range in the 
highest quartile of values.  The chart indicates a number of outliers above 40.

```{r}
ggplot(data=wine, aes(x=total.sulfur.dioxide)) +
  geom_histogram(binwidth = 5)+
  scale_x_continuous(breaks = seq(0, 300, 10))
summary(wine$total.sulfur.dioxide)
```

This histogram for total sulfur dioxide shows a positively skewed distribution. 
There are a few extreme outliers over 280.

```{r}
ggplot(data=wine, aes(x=sulphates)) +
  geom_area(stat='bin') +
  scale_x_continuous(breaks = seq(0, 2, .1))
summary(wine$sulphates)
```

This sulphates area plot shows the majority of the values are between 0.5 and
0.9, with a few over 1.5 g/dm^3.  

```{r}
ggplot(data=wine, aes(x=residual.sugar)) +
  geom_freqpoly(binwidth = .2)+
  scale_x_continuous(breaks = seq(0, 16, .5))
summary(wine$residual.sugar)
```

This frequency polygon for ersidual sugar shows a very long tail, with the 
majority of wines having between 1 and 3 grams/dm^3.  The long tail pulls the
average residual sugar content to almost the third quartile.

```{r}
ggplot(data=wine, aes(x=chlorides)) +
  geom_freqpoly(binwidth = .005)+
  scale_x_continuous(breaks = seq(0, 1, .05))
summary(wine$chlorides)
```

This chlorides frequency polygon also shows a very long tail.  The majority of 
the chlorides in the wines is between 0.05 and 0.10.  Similar to residual sugar,
the mean value is almost at the third quartile.

```{r}
ggplot(data=wine, aes(x=density)) +
  geom_histogram(binwidth = .0001)+
  scale_x_continuous(breaks = seq(0, 2, .001))
summary(wine$density)
```

This histogram for the density of the wines shows a normal distribution.

```{r}
ggplot(data=wine, aes(x=pH)) +
  geom_histogram(binwidth = .01)+
  scale_x_continuous(breaks = seq(0, 5, .1))
summary(wine$pH)
```

This histogram showing the pH of the wines shows a normal distribution, with a 
few outliers aaround 2.75, 3.9, and 4.0

```{r}
ggplot(data=wine, aes(x=alcohol)) +
  geom_histogram(binwidth = .1)+
  scale_x_continuous(breaks = seq(8, 15, 1))
summary(wine$alcohol)
```

The histogram for alcohol is positively skewed, with half of the values being 
between 9.5 and 11.10%.



# Univariate Analysis

**Dataset Structure**

This dataset is composed of 1599 red wines with 12 features: fixed acidity, 
volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, 
total sulfur dioxide, density, pH, sulphates, alcohol, and quality.  With the
exception of quality, all of the features reflect the chemical composition of 
the wine, and the values are stored numerically.  The quality variable was 
determined subjectively by a panel of three wine experts, with the median score
being taken.  This value is stored as an integer.

There is a variable named 'X', which is an index number for the wine.  This 
variable has been removed from the dataset.


**Main Features**

The main feature of interest within the dataset is the quality.  This is because
it is the only subjective variable, and is determined by the aroma and taste of 
the wine.  The other features are properties that alter the smell and taste, 
which can affect the quality grade.

**Supporting Features**

I think that the acidity values (fixed, volatile, and citric) and the sulfur 
values (free sulfur dioxide, total sulfur dioxide, and sulphates) will support 
my investigation into the quality the most since those will impact the smell 
and taste of the wines the most.

**New Variables**

The total.acid variable was created which is the sum of fixed acid, volatile 
acid, and citric acid.  This is because I want to know if any particular acid 
affects the quality of the wine, or if it is the total amount of acid within 
the wine, regardless of type.  

**Plot Adjustments**

A univariate examination was conducted on all features, and no operations were 
performed to adjust the data.  This is because I wanted to see whe whole set of
data, including the outliers, since the irregularities could contribute to the
overall quality of the wine.  

# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
# Create a correlation matrix and graph it on a corrplot
cor <- cor(wine)
corrplot(cor, method="circle", type='lower', order='hclust')
```

From the correlation plot above, a few things stand out.  The negative 
correlation between the acids and pH makes sense, because more acid will result
in a lower pH.  The positve correlation between free sulfur dioxide and total 
sulfur dioxide also makes sense, because free sulfur dioxide is a component of
total sulfur dioxide.

Lets dive deeper into the correlations between quality and alcohol, volatile 
acidity, citric acid, and sulphates.

```{r}
# create a box plot of quality and alcohol.  Convert quality into a factor
# to have multiple box plots by quality.

ggplot(data=wine, aes(x=as.factor(quality), y=alcohol)) +
  geom_boxplot() +
  stat_summary(fun.y='mean', geom='point', color='red')

ggplot(data=wine, aes(x=alcohol)) +
  geom_density(aes(color=as.factor(quality)))
```

These plots look at alcohol content compared to quality.  The first chart is a 
boxplot showing the central tendency of alcohol for the different quality 
grades.  The second chart shows the kernel density of alcohol by quality.  It 
appears that higher quality wines have a higher alcohol content.  

```{r}
# Plotting quality (as a factor) against volatile acidity
ggplot(data=wine, aes(x=as.factor(quality), y=volatile.acidity)) +
  geom_boxplot() +
  stat_summary(fun.y='mean', geom='point', color='red')
by(wine$volatile.acidity, wine$quality, summary)
```

This trend reflects an increasing quality with decreasing volatile acidity.

```{r}
# plotting quality (as a factor) against citric acid
ggplot(data=wine, aes(x=as.factor(quality), y=citric.acid)) +
  geom_boxplot() +
  stat_summary(fun.y='mean', geom='point', color='red')
by(wine$citric.acid, wine$quality, summary)
```

This trend reflects increasing quality with increasing amounts of citric acid. 
It is interesting that for wines of quality 3 and 4, the mean is substantially 
different from the median values.

```{r}
# plotting quality (as a factor) against sulphates
ggplot(data=wine, aes(x=as.factor(quality), y=sulphates)) +
  geom_boxplot() +
  stat_summary(fun.y='mean', geom='point', color='red')
by(wine$sulphates, wine$quality, summary)
```

These box plots show an increase in quality with an increase in sulphates.  An
interesting point to note is how small the IQR is compared to the overall
range of the dataset.  Ther is also a large number of outliers for wines of 
quality 4, 5, and 6.  For wines of quality 4, the outliers are so large that 
it skews the average sulphate amount to almost the third quartile.

```{r}
# plotting volatile acidity and citric acid on as scatter plot
# with a smooth trendline

ggplot(data=wine, aes(x=volatile.acidity, y=citric.acid)) +
  geom_point(alpha=1/5)

ggplot(data=wine, aes(x=volatile.acidity, y=citric.acid)) +
  geom_point(alpha=1/5) +
  scale_x_continuous(limits = c(0, quantile(wine$volatile.acidity, 0.99))) +
  scale_y_continuous(limits = c(0, quantile(wine$citric.acid, 0.99))) +
  geom_smooth(method = 'lm') 
```

This scatter plot looks at the relationship between volatile acidity and citric 
acid.  As volatile acidity increases, citric acid decreases.  The trendline 
shows the negative relationship.

```{r}
# plotting the relationships of sulphates and chlorides, and residual sugar and density
ggplot(data=wine, aes(x=chlorides, y=sulphates)) +
  geom_point(alpha = 1/8) +
  scale_x_continuous(limits = c(0, quantile(wine$chlorides, 0.99))) +
  scale_y_continuous(limits = c(0, quantile(wine$sulphates, 0.99)))

ggplot(data=wine, aes(x=residual.sugar, y=density)) +
  geom_point(alpha = 1/8) +
  scale_x_continuous(limits = c(0, quantile(wine$residual.sugar, 0.99)))
```

As the corrplot above indicates there may be a mild correlation between 
chlorides and sulphates, and between residual sugar and density, the above
pairs were plotted to see the relationships.  While the majority of the data 
is clustered together, there does not appear to be a correlation between the 
features.


```{r}
# Comparing the ratio of citric acid and volatile acidity to the quality
# of the wine

ggplot(data=wine, aes(x=as.factor(quality), y=citric.acid/
                        volatile.acidity)) +
  geom_jitter(alpha = 1/4) +
  geom_hline(yintercept = 1, color='blue') +
  stat_summary(fun.y='mean', geom='line', aes(group=1), color='red', 
               linetype=2) +
  stat_summary(fun.y='median', geom='line', aes(group=2), color='orange', 
               linetype=2)
  
# create a summary table of the above stat_summaries using dplyr
wine.citric_by_volatile <- wine %>%
  group_by(quality) %>%
  summarise(mean_ratio = mean(citric.acid/volatile.acidity), 
            median_ratio =median(citric.acid/volatile.acidity)) %>%
  arrange(quality)

```

This plot looks at the ratio of citric acid and volatile acidity and compares 
it to the quality of the wine.  It looks like as the ratio goes up, the quality 
also goes up, however this is not reflected in the highest quality rated wines.
The red line shows the average ratios for each quality grade, and the orange 
line shows the median ratio for each qualilty grade.


# Bivariate Analysis

**Relationships with Quality**

The quality of the wine was correlated the most with alcohol content at 0.48.  
As the alcohol content increased in the wine, the quality increased as well. 
The average alcohol content of the worst rated wines was 9.955%, and the average
alcohol content of the hgiehst rating wines was 12.09%.  

Volatile acidity had an inverse correlation with quality, -0.39, with the lowest
quality wines having a volatile acidity average of .8845.  The highest quality 
wines had a volatile acidity of .4233.  This relationship is expected, as 
volatile acid is defined as "the amount of acetic acid in wine, which at too 
high of levels can lead to an unpleasant, vinegar taste."

Citric Acid has a small correlation with quality, with a correlation coefficient
of 0.23.  The median value for the worst quality wines was 0.035, and the median 
value for the highest quality wines was 0.42.  This relationship is expected, as
citric acid is described as "found in small quantities, citric acid can add 
'freshness' and flavor to wines." 

Sulphates also has a small positvie correlation with quality, with a correlation 
coefficient of 0.25.  The median value for the worst quality wines was 0.5450, 
and the median value for the highest quality wines was 0.74.  Sulphates are 
described as "a wine additive which can contribute to sulfur dioxide gas (S02) 
levels, wich acts as an antimicrobial and antioxidant." 


**Interesting Relationships**

I thought it was interesting how Citric Acid and Volatile Acidity are inversely
correlated.  In addition to the correlation between Citric Acid and Quality and 
Volatile Acidity and Quality, I also looked at the ratio of Citric Acid and 
Volatile Acidity to Quality.  Because increasing Citric Acid was correlated to 
an increase in Quality, but an increase in Volatile Acidity was negatively 
correlated to Quality, I was curious if a higher Citric Acid to Volatile Acidity
ratio was correlated with an increase in quality.

The majority of wines in the dataset have more volatile acidity than citric 
acid, but as the ratio increases, the quality goes up.  The mean and median 
wines for the highest quality ratings (7 and 8), have a ratio greater than 1.

**Strongest Correlations**

  * -0.68	fixed acidity and pH
  * 0.67	fixed acidity and citric acid
  * 0.67	fixed acidity and density
  * 0.67	free sulfur dioxide and total sulfur dioxide

Because free sulfur dioxide is included in total sulfur dioxide, this strong 
relationship is expected.  


# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

# plotting volatile acidity and citric acid, colored by quality
ggplot(data=wine, aes(x=volatile.acidity, y=citric.acid, color=factor(quality))) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE) +
  scale_color_brewer(type='div', palette='RdYlGn')
```

Here's another comparison of citric acid to volatile acidity, colored by 
quality.

```{r}
# plotting fixed acidity and density, colored by quality
ggplot(data=wine, aes(x=fixed.acidity, y=density, color=factor(quality))) +
  geom_point(alpha = .5) +
  geom_smooth(method='lm', se=FALSE) +
  scale_color_brewer(type='div', palette = 'Accent')
```

As fixed acidity increases, density increases as well.  It also appears that 
quality is impacted by this relationship, as the least dense wines within a 
specific level of fixed acidity are given higher quality ratings.

```{r}
# plotting fixed acidity and ph, colored by citric acid
ggplot(data=wine, aes(x=fixed.acidity, y=pH)) +
  geom_point(alpha = 1/2, aes(color=citric.acid)) +
  scale_color_continuous(low="light blue", high='dark green')

# facet wrapped by quality
ggplot(data=wine, aes(x=fixed.acidity, y=pH)) +
  geom_point(alpha = 1/2, aes(color=citric.acid)) +
  scale_color_continuous(low="light blue", high='dark green') +
  facet_wrap(~quality) +
  geom_smooth(method='lm', color='red', se=FALSE)

```

The previous analysis pointed out an interesting correlation between fixed 
acidity and pH, and density.  Here these three features are plotted together.
As fixed acidity increases, pH decreases, and density increases.

```{r}
# plotting alcohol and citric acid, colored by qualilty
ggplot(data=wine, aes(x=alcohol, y=citric.acid, color=factor(quality))) +
  geom_jitter(alpha = 1/2) +
  scale_color_brewer(type='qual', palette='Set2') +
  geom_smooth(method='lm', se=FALSE)
```

Here we look at how citric acid and alcohol are related to quality.  As quality
increases, the alcohol content increases.  Higher quality wines also appear to
have a higher amount of citric acid, however a wide range of citric acid is used
in wines with a quality of 5 or 6.  It also appears that as alcohol content 
increases, the amount of citric acid increases as well.

```{r}
# plotting total and free sulfur dioxide, colored by residual sugar
# excluding the top 1% values for each sulfur dioxide.

ggplot(data=wine, aes(x=total.sulfur.dioxide, y=free.sulfur.dioxide)) +
  geom_jitter(alpha =1/2, aes(color=residual.sugar)) +
  scale_x_continuous(limits = c(0, quantile(wine$total.sulfur.dioxide, 0.99))) +
  scale_y_continuous(limits = c(0, quantile(wine$free.sulfur.dioxide, 0.99))) +
  scale_color_continuous(low="light blue", high='black') +
  geom_smooth(method='lm', se=FALSE)
```

This plot looks at free sulfur dioxide, total sulfur dioxide, and the levels of 
residual sugar.  Free sulfur dioxide is a component of total sulfur dioxide, and
the positive correlation is apparent.  When colored by residual sugar, it 
appears that the wines with higher amounts of residual sugar contain more total
sulfur dioxide and free sulfur dioxide.

# Multivariate Analysis

**Relationships with Quality**
In the Bivariate analysis, I explored the relationship between citric acid and 
volatile acid and how they impact quality.  In this multivariate analysis, I 
explored this even further.

The highest quality wines not only have a higher citric acid to volatile acidity
ratio, but also a lower density to fixed acidity ratio as well.

Additionally, while citric acid and alcohol content within the wines did not 
appear to have any correlation, it became apparent that citric acid and alcohol
are both significant factors when comparing between quality.  On the last chart,
it shows that higher quality wines not noly have the higher alcohol content, 
but higher amounts of citric acid as well.

**Other Relationships**
From the bivariate analysis, the fixed acidity had a strong correlation 
coefficient with pH, citric acid, and density, so I looked into these 
relationships a bit more.

As fixed acidity increased, pH decreased and citric acid increased.  When 
faceted by quality, it didn't appear that the fixed acidity and pH had any
substantial influence on the quality.  This is further substantiated by the
correlation grid from the Bivariate analysis.

------

# Final Plots and Summary

### Alcohol by Quality
```{r echo=FALSE, Plot_One}
ggplot(data=wine, aes(x=alcohol)) +
  geom_density(aes(color=as.factor(quality))) +
  ggtitle("Kernel Density Estimate of Alcohol Content by Quality") +  
  xlab("Alcohol by Volume (%)") +
  ylab("Gaussian Kernel Density")

```

Plot One looks at the relationship between Alcohol by Volume and Quality.  As
alcohol content was the factor that had the strongest correlation with quality,
this plot shows the kernel density of alcohol volume, colored by quality. As 
you can see in the plot, the lowest quality wines have the highest kernel
density under 11% Alcohol by Volume, while the highest quality wines have the
highest kernel density over 11%.  This shows a trend that more high quality 
wines have higher amounts of alcohol.

### Fixed Acidity and Density
```{r echo=FALSE, Plot_Two}
ggplot(data=wine, aes(x=fixed.acidity, y=density, color=factor(quality))) +
  geom_point(alpha = .5) +
  geom_smooth(method='lm', se=FALSE) +
  scale_color_brewer(type='div', palette = 'Accent', name="Quality") +
  ggtitle("Comparison of Fixed Acidity and Density") +
  xlab("Fixed Acidity (gm/dm^3)") +
  ylab("Density (gm/cm^3)")
```

This plot looks at the relationship of Density and Fixed Acidity, and colors it
by quality.  Fixed acidity and density have a positive correlation, and when 
looking at the wine quality on the plot, an interesting observation is made: 
while the the trendlines for each quality level appears to have roughly the same
slope, the y-intercept decreases with quality. This means that higher quality 
wines are less dense when compared to other wines with the same amounts of fixed 
acidity.

### Impact of Citric Acid and Volatile Acidity on Quality
```{r echo=FALSE, Plot_Three}


# plot the ratio of citric acid to volatile acidity.  Because the summary data 
# is stored in another dataset, the summary lines will have 'x' set to 
# "quality-2" to ensure it lines up properly with the wine dataset.  Labels are
# also added for description.

ggplot(data=wine, aes(x=as.factor(quality), y=citric.acid/
                        volatile.acidity)) +
  geom_jitter(alpha = 1/4) +
  geom_hline(yintercept = 1, color='blue') +
  geom_line(data = wine.citric_by_volatile, aes(x=quality-2, y=mean_ratio, 
                                                color='mean')) +
  geom_line(data = wine.citric_by_volatile, aes(x=quality-2, y=median_ratio, 
                                                color='median')) +
  scale_color_manual('Legend', values = c('mean'='red', 'median'='orange' )) +
  ggtitle("Ratio of Citric Acid and Volatile Acidity compared to Wine Quality")+
  xlab("Quality") +
  ylab("Ratio of Citric Acid to Volatile Acidity")

```

This plot looks at the ratio of Citric Acid and Volatile Acidity and compares 
them by quality.  the mean and median ratios are plotted on the above table in
red and orange respectively.  This diagram shows that as the ratio increases, 
the quality of the wine increases as well.  In fact, of the top quality wines, 
rated 7 and 8, the mean and median wines have a ratio greater than 1.  This 
means higher quality wines contain more citric acid than volatile acidity.


# Reflection

This red wine dataset contains information on 1599 different wines, with 11 
physical features and 1 subjective feature.  This subjective feature is quality,
and was determined by a panel of three wine experts.  To explore this dataset, 
one question stood out: What determines the quality of each wine?

From the initial examination of data, I thought that acids, sulfurs, and 
sulphates would be the biggest contributor towards wine quality.  However the 
feature that correlated the highest with quality was neither of them:  it was 
alcohol content.  While larger amounts of alcohol doesn't cause higher quality,
wines with larger alcohol by volume amounts tended to be rated higher by the 
wine experts.

Citric acid and volatile acidity also were correlated with the quality of the
wines as well:  Citric acid had a positive correlation, while volatile acid had 
a negative correlation.  Their inverse relationship with quality interested me,
so I not only investigated the relationship between the two acids, but the ratio
of the two when compared to quality as well.  This lead to the observation that 
wines with a higher citric acid to volatile acidity ratio tend to have a higher
quality.

Another interesting observation from the dataset is how fixed acidity has a 
strong correlation with pH and citric acid.  Hoewver when a plot was faceted by 
quality, it didn't appear that fixed acidity had a correlation with quality.  
But when fixed acidity was plotted with density, it appears that higher quality
wines have lower densities compared to other wines with the same amount of fixed
acidity.

Because this dataset is mostly composed of the chemical composition of these 
wines, there are a lot of limitations on this analysis.  The length of 
fermentation, types of grapes used, type of container the wine was fermented in
(wood or plastic?  what type of wood?), and other additives contribute to the 
quality of wine, but were not included in the dataset.  Additionally, it would 
be interesting to know the year of some wines, to see if the wines maintain the
same chemical properties every year, or what values change.
