---
title: 'Lesson 6: Problem Set Explore 2 Variables'
output:
  html_document: default
  html_notebook: default
---

1. Quiz: price vs. x

```{r}
library(ggplot2)
data(diamonds)

ggplot(data=diamonds, aes(x=x, y= price)) +
  geom_point()
```


3. Quiz: Correlations
```{r}
cor.test(diamonds$price, diamonds$x)
cor.test(diamonds$price, diamonds$y)
cor.test(diamonds$price, diamonds$z)
```

4. Quiz: price vs. depth
```{r}
ggplot(data = diamonds, aes(x = depth, y = price)) + 
  geom_point()
```

5: Quiz - Adjustments price vs depth
# Change the code to make the transparency of the
# points to be 1/100 of what they are now and mark
# the x-axis every 2 units.

```{r}
ggplot(data = diamonds, aes(x = depth, y = price)) + 
  geom_point(alpha= 1/100) +
  scale_x_continuous(breaks =seq(43,79,2))
```

7. Quiz: Correlation - price and depth
```{r}
cor.test(diamonds$depth, diamonds$price)
```

8. Quiz - price vs. carat
# Create a scatterplot of price vs carat
# and omit the top 1% of price and carat
# values.

```{r}
#quantile(diamonds$carat, 0.99)

diamonds.sub <- subset(diamonds, carat < quantile(diamonds$carat, 0.99) | 
                         price < quantile(diamonds$price, 0.99))
ggplot(data=diamonds.sub, aes(x=carat, y=price)) +
  geom_point()
```


9: Quiz - price vs. volume
```{r}
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
ggplot(data = diamonds, aes(x=volume, y=price)) +
  geom_point()
```

11. Quiz - Correlations on Subsets
```{r}
diamonds.corsub <- subset(diamonds, volume > 0 & volume <= 800)
cor.test(diamonds.corsub$price, diamonds.corsub$volume)
```

12. Quiz: Adjustments - price vs volume
```{r}
ggplot(data=diamonds.corsub, aes(x=volume, y=price)) +
  geom_point(alpha = 1/50) +
  geom_smooth()
```


13. Quiz: Mean Price by Clarity
# Use the function dplyr package
# to create a new data frame containing
# info on diamonds by clarity.

# Name the data frame diamondsByClarity

# The data frame should contain the following
# variables in this order.

#       (1) mean_price
#       (2) median_price
#       (3) min_price
#       (4) max_price
#       (5) n

```{r}
library(dplyr)
diamondsByClarity <- diamonds %>%
  group_by(clarity) %>%
  summarise(mean_price = mean(price),
            median_price = median(price),
            min_price = min(price),
            max_price = max(price),
            n = n())

head(diamondsByClarity)
            
```


14. Quiz: Bar Charts of Mean Price
```{r}
library(gridExtra)
diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, 
                                    mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, 
                                  mean_price = mean(price))

d_by_clarity <- ggplot(data=diamonds_mp_by_clarity, aes(x=clarity, y=mean_price)) +
  geom_bar(stat='identity')


d_by_color <- ggplot(data=diamonds_mp_by_color, aes(x=color, y=mean_price)) +
  geom_bar(stat='identity')

grid.arrange(d_by_clarity, d_by_color)

```



15. Quiz: Trends in Mean Price

mean price decreases as clarity improves
mean price increases as color gets worse


















