---
title: "Problem Set 8"
output:
  html_document: default
  html_notebook: default
---
```{r echo=FALSE}
library(ggplot2)
```


1. Quiz: Price Histograms with Facet and Color
```{r Price Histograms with Facet and Color}
data(diamonds)
ggplot(data=diamonds, aes(x=price)) +
  geom_histogram(aes(fill=cut), binwidth=100) +
  facet_wrap(~color, scales='free') +
  scale_fill_brewer(type='qual')
  
```



2. Quiz: Price vs. Table Colored by Cut
```{r Price vs. Table Colored by Cut}
ggplot(data=diamonds, aes(x=table, y=price)) +
  geom_point(aes(color=cut))
```


3. Quiz: Typical Table Value
Ideal Cut has a table range of 53 to 57
Premium Cut has at able range of 58 to 62

4. Quiz: Price vs. Volume and Diamond Clarity
```{r Price vs. Volume and Diamond Clarity}
diamonds$volume = diamonds$x * diamonds$y * diamonds$z

ggplot(data=diamonds, aes(x=volume, y=price)) +
  geom_point(aes(color=clarity)) +
  scale_y_log10() +
  scale_color_brewer(type = 'div') +
  xlim(0, quantile(diamonds$volume, 0.99))
```


5. Quiz: Proportion of Friendships Initiated
```{r Proportion of Friendships Initiated}
pf <- read.csv('pseudo_facebook.tsv', sep='\t')
pf$prop_initiated <- pf$friendships_initiated / pf$friend_count
```


6. Quiz: prop_initiated vs. tenure
```{r prop_initiated vs. tenure}
pf$year_joined <- floor(2014 - (pf$tenure/365))
pf$year_joined.bucket <- cut(pf$year_joined,
                             c(2004, 2009, 2011, 2012, 2014))

ggplot(data=pf, aes(x=tenure, y=prop_initiated)) +
  geom_line(aes(color=year_joined.bucket), stat = 'summary', fun.y = median)
```


7. Quiz: Smoothing prop_initiated vs tenure
```{r Smoothing prop_initiated vs tenure}
ggplot(data=pf, aes(x=tenure, y=prop_initiated)) +
  geom_line(aes(color=year_joined.bucket), stat = 'summary', fun.y = median) +
  geom_smooth()
```


8. Quiz: Greatest prop_initiated Group
Users who joined after 2012 have the highest proportion of friends initiated


9. Quiz: Largest Group Mean prop_initiated 
```{r Largest Group Mean prop_initiated }
by(pf$prop_initiated, pf$year_joined.bucket, summary)
```


10. Quiz: Price/Carat Binned, Faceted and Colored
```{r Price/Carat Binned, Faceted and Colored}
ggplot(data=diamonds, aes(x=cut, y=(price/carat))) +
  geom_jitter(aes(color=color)) +
  facet_wrap(~clarity) +
  scale_color_brewer(type='div')
```

