---
title: "R Notebook"
output: html_notebook
--- 


Import the diamonds data set and print a summary of the information
```{r}
library(ggplot2)
data(diamonds)
summary(diamonds)
```


Create a histogram of all diamond prices
```{r}
qplot(data = diamonds, x = price)
```


Finding number of diamonds that cost less than 250, less than 500, and over 15000.
```{r}
p <- diamonds['price']
sum (p < 250)
sum (p < 500)
sum (p >= 15000)
```


Price analysis of the peak of the histogram
```{r}
qplot(data = diamonds, x = price, binwidth = 10, color = I('black')) +
  scale_x_continuous(lim=c(0,2000), breaks=seq(0,2000,100))
```

Price analysis of diamonds by cut
```{r}
qplot(data = diamonds, x = price) +
  facet_wrap (~cut)
```

Summary statistics of diamonds by cut
```{r}
by(diamonds$price, diamonds$cut, summary)
```

Modifying facet wrap histograms with variable y-axis
Set scales=free 
```{r}
qplot(x = price, data = diamonds) + facet_wrap(~cut, scales = "free")
```

Histogram of price per carat, by cut.
x-axis transformation using log10
```{r}
qplot(x = log10(price/carat), data = diamonds, binwidth = .05) + 
  facet_wrap(~cut, scales = "free")
```


Boxplot and summary of price by color
```{r}
qplot(data = diamonds, geom = "boxplot", x = color, y = price)
by(diamonds$price, diamonds$color, summary)
```

IQR of Diamonds by specific color
```{r}
IQR(subset(diamonds, color == 'J')$price)
```

Boxplots showing price per carat of diamonds across colors
```{r}
qplot(data = diamonds, geom = "boxplot", x = color, y = (price/carat))
```

Frequency polygon showing weight of diamonds
```{r}
qplot(data = diamonds, geom='freqpoly', x=carat, binwidth=.01) +
  scale_x_continuous(breaks = seq(0,5, .1))
```



